---
title: "Is my judge any good?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Is my judge any good?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(1)
```

```{r setup}
library(yourhonor)
```

When you score model output with an LLM-as-judge, you inherit a question most
eval tooling quietly skips: **is the judge any good?** A judge that disagrees
with humans, drifts run to run, or quietly rewards longer answers will hand you
confident numbers that mean nothing. `yourhonor` treats the judge as a measurement
instrument and reports its psychometrics, so you can say *"this judge agrees with
humans at AC1 = 0.81, shows no significant position bias, and flips on 4% of
re-runs"* instead of *"we used an LLM judge"*.

`yourhonor` does not run evaluations — `vitals` does that. It measures the judge
that `vitals` (or any other harness) uses to score them.

## The judge contract

Everything in `yourhonor` talks to a judge through one small contract: a function
that takes a data frame of items and returns one label per row.

```{r}
# A toy judge for the vignette: it agrees with the reference answer most of the
# time but is a little lenient. In real use this wraps an ellmer chat (see below).
toy_judge <- function(data) {
  ref <- data$reference
  ifelse(runif(nrow(data)) < 0.85, ref,
         ifelse(ref == "P", "C", ref))
}
```

We also need a **gold set**: items a human has labeled. `yourhonor` ships a small
synthetic one (`support_gold`); here we build an inline example so the vignette
is self-contained.

```{r}
gold <- tibble::tibble(
  input = paste("Question/answer pair", 1:60),
  target = "Does the answer resolve the question?",
  reference = sample(c("C", "P", "I"), 60, replace = TRUE, prob = c(.6, .25, .15)),
  human_label = NA_character_
)
# treat the reference as the human's verdict for this demo
gold$human_label <- gold$reference
```

## Calibrate against humans

`calibrate_judge()` runs the judge over the gold set and reports chance-corrected
agreement with humans, with bootstrap confidence intervals.

```{r}
cal <- calibrate_judge(gold, toy_judge, n_boot = 300, seed = 1)
cal
```

The coefficients are reported together on purpose. Cohen's and Fleiss' kappa are
the familiar baseline; Gwet's AC1 is robust when one label dominates (as
"Correct" does here); Krippendorff's alpha generalizes across scales. When AC1
sits noticeably above kappa, your labels are skewed and kappa is *understating*
real agreement.

```{r}
agreement(cal)
```

## See *where* it disagrees

An aggregate number hides the shape of the error. `confusion()` shows whether a
judge fails good answers or over-credits weak ones.

```{r}
confusion(cal, normalize = "row")
```

```{r, eval = requireNamespace("ggplot2", quietly = TRUE), fig.width = 5, fig.height = 4}
plot_confusion(cal)
```

## Against more than one human

If several people labeled the gold set, pass every rater column. `yourhonor`
compares the judge to the human *consensus* (per-item majority) and reports the
human-human **ceiling** — the judge can't be expected to agree with people more
than they agree with each other.

```{r}
multi <- tibble::tibble(
  human_label   = c("C", "C", "P", "I", "C", "P", "I", "C"),
  human_label_2 = c("C", "P", "P", "I", "C", "C", "I", "C"),
  human_label_3 = c("C", "C", "P", "P", "C", "P", "I", "C"),
  judge_label   = c("C", "C", "C", "I", "C", "P", "P", "C")
)
calibrate_judge(
  multi,
  human_cols = c("human_label", "human_label_2", "human_label_3"),
  n_boot = 200, seed = 1
)
```

## Probe for bias

`probe_bias()` perturbs each item in a way a fair judge should ignore, then
measures how often — and in which direction — the verdict flips. Here we use a
deliberately length-sensitive judge to show verbosity bias being caught.

```{r}
length_judge <- function(data) ifelse(nchar(data$input) > 25, "C", "P")
probe_bias(gold, length_judge,
           probes = c("verbosity", "self_preference", "bandwagon"), seed = 1)
```

A significant flip rate in the `favors_perturbed` direction on the `verbosity`
probe is the signal that your judge is rewarding length, not quality.

## Check reproducibility

A judge whose verdicts wobble between identical runs is not yet an instrument.
`test_retest()` quantifies that wobble.

```{r}
noisy_judge <- function(data) {
  ifelse(runif(nrow(data)) < 0.9, data$reference, "C")
}
test_retest(gold, noisy_judge, n_runs = 6, seed = 1)
```

## One-paragraph verdict

`report()` rolls calibration, bias and stability into a single call you can paste
beside an eval or into a README.

```{r}
bias <- probe_bias(gold, length_judge, probes = "verbosity", seed = 1)
stab <- test_retest(gold, noisy_judge, n_runs = 6, seed = 1)
report(cal, bias = bias, stability = stab, output = "console")
```

## Wiring a real judge (ellmer / vitals)

In production your judge is a model. The contract is the same: wrap the call in a
function that returns one label per row. This chunk is not run here because it
needs an API key.

```{r, eval = FALSE}
library(ellmer)

# A judge backed by a real model. Swap in your own grading prompt.
model_judge <- function(data) {
  chat <- chat_anthropic(
    system_prompt = "Grade the answer as exactly one of C, P, or I.",
    model = "claude-sonnet-4-6"
  )
  vapply(seq_len(nrow(data)), function(i) {
    chat$clone()$chat(paste0(data$target[i], "\n\n", data$input[i]))
  }, character(1))
}

calibrate_judge(support_gold, model_judge, seed = 1)
```

If you already grade with a `vitals` model-graded scorer, wrap that scorer the
same way — call it inside the judge function and return its C/P/I grades — and
`yourhonor` will calibrate it without changing your eval.
